European or American? A re-examination of the transatlantic influence over the IMF

### 2025-09-11 manuscript/manuscript.qmd
Improved keywords and added comprehensive conclusion section
Co-authored-by: kv9898 <105025148+kv9898@users.noreply.github.com>

---
title: European or American? A re-examination of the transatlantic influence over the IMF
date: last-modified
date-format: |
    | DD MMMM YYYY
authors:
  - name: Dianyi Yang
    email: dianyi.yang@politics.ox.ac.uk
    orcid: 0009-0004-4652-3429
    corresponding: true
    affiliations:
      - ref: ox    
  - name: Tong Hu
    orcid: 0009-0003-5234-8164
    affiliations:
      - ref: bw
    equal-contributor: true
  - name: Zhenyi Chen
    orcid: 0009-0008-2910-5187
    affiliations:
      - ref: rmu
    equal-contributor: true
affiliations:
  - id: ox
    name: Department of Politics and International Relations, University of Oxford
  - id: bw
    name: School of International Relations and Diplomacy, Beijing Foreign Studies University
  - id: rmu
    name: School of International Studies, Renmin University of China
format:
  #native: default
  #docx: default
  orcid-pdf:
    keep-tex: true
    number-sections: true
    include-in-header:
      - text: |
          \newcommand{\centerstart}{\begin{center}}
          \newcommand{\centerend}{\end{center}}
          \usepackage{amsthm, thm-restate}
          \newtheorem{hyp}{Hypothesis}
          \newtheorem{subhyp}{Hypothesis}[hyp]
          \renewcommand{\thesubhyp}{\thehyp\alph{subhyp}}
          \usepackage{chngcntr}
          \usepackage{apptools}
          \AtAppendix{\counterwithin{figure}{section}}
          \AtAppendix{\counterwithin{table}{section}}
abstract: "This paper investigates whether the United States or Europe holds greater influence over the International Monetary Fund (IMF)—a longstanding yet underexamined question in the study of global economic governance. While the Fund is widely regarded as dominated by “Western” powers, most existing research either focuses exclusively on American influence or treats the West as a unified bloc, masking important variation between its two key actors. Drawing on panel data from 1980 to 2010, we construct separate measures of US and European influence using principal component analysis of commonly used alignment indicators—UN voting patterns, bilateral trade, and banking exposure—and estimate a series of Tobit and Probit models to assess their impact on four key IMF lending outcomes: loan-to-GDP ratios, participation rates, loan approvals, and conditionality. We find that European influence is strongly associated with higher loan volumes, more frequent participation, and greater likelihood of loan approval, while US influence is not. However, American ties are significantly correlated with fewer conditions attached to loans. Additional analysis shows that European influence is not driven by regional or colonial favouritism: neither European nor African borrowers benefit more from European ties—and may, in fact, fare worse—than other countries with comparable alignment. These findings complicate existing assumptions about IMF decision-making, highlighting the need to distinguish between types of Western influence. They also suggest that Europe wields broader institutional influence over access to IMF resources, while US power manifests more selectively through conditionality. In doing so, the paper contributes to debates on hegemony, institutional legitimacy, and intra-Western dynamics in global economic governance."
keywords: IMF, US hegemony, Europe
execute:
  echo: false
  warning: false
  error: false
  message: false
---

@@

# Case Study

@@

>>>>

---
title: European or American? A re-examination of the transatlantic influence over the IMF
date: last-modified
date-format: |
    | DD MMMM YYYY
authors:
  - name: Dianyi Yang
    email: dianyi.yang@politics.ox.ac.uk
    orcid: 0009-0004-4652-3429
    corresponding: true
    affiliations:
      - ref: ox    
  - name: Tong Hu
    orcid: 0009-0003-5234-8164
    affiliations:
      - ref: bw
    equal-contributor: true
  - name: Zhenyi Chen
    orcid: 0009-0008-2910-5187
    affiliations:
      - ref: rmu
    equal-contributor: true
affiliations:
  - id: ox
    name: Department of Politics and International Relations, University of Oxford
  - id: bw
    name: School of International Relations and Diplomacy, Beijing Foreign Studies University
  - id: rmu
    name: School of International Studies, Renmin University of China
format:
  #native: default
  #docx: default
  orcid-pdf:
    keep-tex: true
    number-sections: true
    include-in-header:
      - text: |
          \newcommand{\centerstart}{\begin{center}}
          \newcommand{\centerend}{\end{center}}
          \usepackage{amsthm, thm-restate}
          \newtheorem{hyp}{Hypothesis}
          \newtheorem{subhyp}{Hypothesis}[hyp]
          \renewcommand{\thesubhyp}{\thehyp\alph{subhyp}}
          \usepackage{chngcntr}
          \usepackage{apptools}
          \AtAppendix{\counterwithin{figure}{section}}
          \AtAppendix{\counterwithin{table}{section}}
abstract: "This paper investigates whether the United States or Europe holds greater influence over the International Monetary Fund (IMF)—a longstanding yet underexamined question in the study of global economic governance. While the Fund is widely regarded as dominated by “Western” powers, most existing research either focuses exclusively on American influence or treats the West as a unified bloc, masking important variation between its two key actors. Drawing on panel data from 1980 to 2010, we construct separate measures of US and European influence using principal component analysis of commonly used alignment indicators—UN voting patterns, bilateral trade, and banking exposure—and estimate a series of Tobit and Probit models to assess their impact on four key IMF lending outcomes: loan-to-GDP ratios, participation rates, loan approvals, and conditionality. We find that European influence is strongly associated with higher loan volumes, more frequent participation, and greater likelihood of loan approval, while US influence is not. However, American ties are significantly correlated with fewer conditions attached to loans. Additional analysis shows that European influence is not driven by regional or colonial favouritism: neither European nor African borrowers benefit more from European ties—and may, in fact, fare worse—than other countries with comparable alignment. These findings complicate existing assumptions about IMF decision-making, highlighting the need to distinguish between types of Western influence. They also suggest that Europe wields broader institutional influence over access to IMF resources, while US power manifests more selectively through conditionality. In doing so, the paper contributes to debates on hegemony, institutional legitimacy, and intra-Western dynamics in global economic governance."
keywords: International Monetary Fund, transatlantic influence, global economic governance, conditionality, principal component analysis, institutional power, financial diplomacy
execute:
  echo: false
  warning: false
  error: false
  message: false
---

@@

@@

This paper has examined a fundamental yet underexplored question in the study of global economic governance: who wields greater influence over the International Monetary Fund—the United States or Europe? By disaggregating "Western" influence into its constituent components and employing principal component analysis to construct separate measures of US and European alignment, we have uncovered important asymmetries in how transatlantic powers exercise their influence within the Fund.

Our empirical analysis yields several key findings that challenge conventional wisdom about IMF decision-making. First, European influence demonstrates a broader and more consistent association with favourable lending outcomes than previously recognised. Countries with stronger ties to major European powers—the United Kingdom, France, and Germany—receive significantly larger loans relative to GDP, participate more frequently in IMF programmes, and enjoy higher approval rates for new lending arrangements. These effects are statistically significant and substantively meaningful, suggesting that European actors wield considerable institutional influence over the Fund's core lending functions.

Second, contrary to much existing literature that emphasises American dominance, US influence shows a more selective pattern of impact. While American ties do not significantly affect loan volumes, participation rates, or approval decisions, they are strongly associated with more lenient conditionality—fewer performance criteria, benchmarks, and prior actions attached to lending arrangements. This finding suggests that US power manifests not through broad institutional control over access to resources, but through targeted influence over the terms and conditions of assistance.

Third, our robustness checks reveal that European influence operates on a global rather than regional or colonial basis. Neither European borrowers nor African recipients benefit disproportionately from their ties to European powers. If anything, the evidence suggests these groups may receive less favourable treatment than other countries with comparable alignment profiles. This pattern indicates that European influence reflects genuine institutional power within the IMF rather than geographic or historical favouritism.

These findings have several important implications for our understanding of international financial institutions and global economic governance more broadly. Theoretically, they complicate simple narratives of American hegemony in international organisations. While the United States undoubtedly remains the most powerful single actor within the Bretton Woods system, our results suggest that European powers collectively exercise substantial influence over key institutional functions. This finding aligns with observations about the IMF's historical leadership structure—epitomised by the "gentlemen's agreement" reserving the Managing Director position for Europeans—and suggests that informal power arrangements may be more consequential than formal voting shares or veto rights.

The asymmetric nature of US and European influence also speaks to broader debates about the sources and exercise of institutional power. European influence appears to operate through channels that affect aggregate resource allocation—determining who receives assistance and in what amounts. American influence, by contrast, manifests through mechanisms that shape the policy content of lending arrangements—affecting what borrowers must do to receive and maintain access to funds. This division of labour may reflect different institutional strengths, strategic priorities, or coordination mechanisms between the two actors.

From a policy perspective, these findings have implications for ongoing debates about IMF reform and representation. Calls for enhanced emerging market voice and reduced Western dominance must grapple with the reality that "the West" is not a monolithic bloc. European and American interests may diverge in important ways, creating opportunities for coalition-building and institutional change that would not exist under unified Western control. Understanding these internal dynamics may be crucial for reform advocates seeking to reshape global financial governance.

Our analysis also sheds light on the nature of conditionality, one of the most controversial aspects of IMF operations. The finding that American influence is associated with fewer conditions, while European influence shows no such effect, suggests that critiques of overly intrusive conditionality may need to be more precisely targeted. European powers, despite their historical association with social democratic governance models, do not appear to advocate for more lenient policy requirements. This pattern may reflect the influence of German-style fiscal conservatism within European policy circles or the adoption of neoliberal economic orthodoxy by European technocratic elites.

Several limitations of our analysis should be acknowledged. First, our sample period ends in 2010, just as the European sovereign debt crisis was beginning to reshape transatlantic dynamics within the IMF. The Fund's involvement in European bailouts through the Troika arrangements may have altered the balance of influence in ways not captured by our data. Future research should examine whether the patterns we document persist in the post-crisis period.

Second, our measures of influence, while grounded in the existing literature, capture alignment rather than direct lobbying or pressure. We cannot observe the specific mechanisms through which influence is exercised or the content of behind-the-scenes negotiations. Qualitative case studies examining specific lending decisions could provide valuable complementary insights into the micro-foundations of the patterns we document.

Third, our focus on the three largest European powers—the UK, France, and Germany—may understate the collective influence of the European Union as a whole. As EU coordination mechanisms develop and potentially consolidate European representation within international financial institutions, the dynamics we analyse may continue to evolve. Research examining the role of EU institutional actors, such as the European Central Bank or European Commission, in IMF decision-making would provide a useful extension of our work.

Fourth, our analysis focuses exclusively on lending outcomes and does not examine other dimensions of IMF operations, such as surveillance, technical assistance, or institutional governance. European and American influence may manifest differently across these various functions. Similarly, our study does not address the effectiveness or appropriateness of the lending arrangements influenced by great power politics—a question that would require additional data on economic outcomes and counterfactual analysis.

Looking ahead, several promising avenues for future research emerge from our findings. First, examining the mechanisms through which European coordination occurs within the IMF would help illuminate the institutional foundations of the influence we document. Studies of EURIMF and SCIMF activities, European Central Bank participation in IMF decisions, or coordination within G7 and G20 forums could provide valuable insights into the operational aspects of European influence.

Second, extending our analysis to include other potential great power influencers—particularly China, whose economic weight and IMF membership have grown substantially since our sample period—would help assess whether the patterns we document reflect specifically transatlantic dynamics or broader features of great power competition within international organisations.

Third, examining variation in influence across different types of IMF programmes—such as Stand-By Arrangements, Extended Fund Facilities, or newer instruments like the Flexible Credit Line—could reveal whether European and American influence operates differently depending on the specific lending modality or the severity of the borrower's economic difficulties.

Finally, our findings suggest the value of disaggregating "Western" influence in studies of other international organisations. The World Bank, regional development banks, and newer institutions like the Asian Infrastructure Investment Bank may exhibit similar asymmetries in great power influence that are masked by treating Western actors as a unified bloc.

In conclusion, this paper demonstrates that the International Monetary Fund, despite its technocratic veneer, remains deeply shaped by geopolitical considerations. However, the nature of great power influence is more complex and differentiated than often assumed. European powers exercise broad institutional influence over access to IMF resources, while American power manifests more selectively through conditionality. This division of influence has important implications for understanding both the current operations of the international financial system and the prospects for its future reform. As emerging economies continue to challenge Western dominance in global governance, recognising the internal dynamics and potential fractures within the Western coalition becomes increasingly important for scholars, policymakers, and reform advocates alike.
